Hospitalizations for psychiatric conditions among adolescents during the COVID-19 pandemic

### 2021-08-18 psy_peds.Rmd
adding percentage plots

all_codes <- obs_raw %>% 
  full_join( clinical_raw,  by = c("patient_num", "cohort", "days_since_admission")) %>%
  full_join( demo_raw, by = c("patient_num", "cohort")) %>%
  dplyr::filter( concept_type == "DIAG-ICD10", 
                 in_hospital == 1,
                 age < 18 & age > 10,
                 date >= history_start_date & date <= history_end_date) %>%
  dplyr::select( patient_num, cohort, sex, age, concept_type, concept_code, calendar_date,
                 in_hospital, n_hospitalisation, len_hospitalisation, date, weeks, year,
                 period, disorder_group, description )

@@

## Diagnoses count per week
Here we calculate the number of patients hospitalized per week (exactly, the number of patients being assigned a diagnosis on that week, hence usually corresponding to the end of hospitalization, ie not taking into account the length of the encounter). 
The barplot shows the weeks on the x-axis, and the patient counts on the y-axis. The horizontal lines is the (quadratic) temporal trend of these patient counts.
```{r}
patient_count_per_week <- input_data[, c("patient_num", "cohort", "weeks")] %>%
  unique() %>%
  group_by(weeks) %>%
  tally() %>%
  mutate(n = ifelse( n < obfuscation | isFALSE(obfuscation),
                     n,
                     0.5),
         siteid = site )

#estimate the number of patients per week
patient_count_per_week %>%
  dplyr::mutate(weeks = as.Date(weeks)) %>% 
  dplyr::mutate(period = ifelse( weeks < pandemic_start_date,
                                  "before_pandemic",
                                 "during_pandemic")) %>% 
  ggplot(aes(x = weeks, y = n, fill = period)) +
  geom_bar(stat="identity")+
  scale_fill_manual(values = cbPalette) +
  labs(y = "Counts",
       x = "Calendar Weeks",
       title = "Weekly patient counts with mental health related ICD codes") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)

ggsave("output/overall_weekly_count_all.png", width = 18, height = 9, units="cm")
```

```{r}
patient_count_per_week <- all_codes %>% 
  mutate(icd_code_category = ifelse(concept_code %in% icdCodes$ICD10_Code, 
                               "psy", 
                               "somatic")
  ) %>% 
  select(patient_num, cohort, weeks, icd_code_category, period) %>% 
  unique() %>%
  group_by(weeks, icd_code_category, period) %>%
  tally() %>%
  mutate(n = ifelse( n < obfuscation | isFALSE(obfuscation),
                     n,
                     0.5),
         siteid = site )

#estimate the number of patients per week
patient_count_per_week %>%
  ggplot(aes(x = weeks, y = n, fill = icd_code_category, color = icd_code_category)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  scale_fill_manual(values = cbPalette) +
  scale_color_manual(values = cbPalette) +
  labs(y = "Counts",
       x = "Calendar Weeks",
       title = "Weekly patient counts with mental health related ICD codes") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)

ggsave("output/overall_weekly_count_all_comparison.png", width = 18, height = 9, units="cm")
```
```{r}
icdCategory <- read.delim("public-data/icd10Codes.txt", header = FALSE, colClasses = "character")
colnames(icdCategory) <- c("icd_code_category", "icd_description_category")
patient_count_per_week_byCategory <- all_codes %>% 
  mutate( icd_code_category = ifelse(substr(concept_code, 1, 1) %in% c("D", "H"), substr(concept_code, 1, 2), substr(concept_code, 1, 1) ))

patient_count_per_week_byCategory <- left_join( patient_count_per_week_byCategory, icdCategory) %>%
  select(patient_num, cohort, weeks, icd_description_category, period) %>% 
  unique() %>%
  group_by(weeks, icd_description_category, period) %>%
  tally() %>%
  mutate(n = ifelse( n < obfuscation | isFALSE(obfuscation),
                     n,
                     0.5),
         siteid = site )

patient_count_per_week_byCategory %>%
  filter( icd_description_category %in% c("Neoplams","Diseases of the respiratory system", "Diseases of the nervous system", "Diseases of the digestive system", "Diseases of the blood", "Congenital malformations, deformations and chromosomal abnormalities" )) %>%
  ggplot(aes(x = weeks, y = n, fill = icd_description_category, color = icd_description_category)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  scale_fill_manual(values = cbPalette) +
  scale_color_manual(values = cbPalette) +
  labs(y = "Counts",
       x = "Calendar Weeks",
       title = "Weekly patient counts") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)

ggsave("output/overall_weekly_count_all_comparison_other_categories.png", width = 18, height = 9, units="cm")

@@

@@

patient_count_per_week_disorder_group <- input_data[input_data$in_hospital == 1, c("patient_num", "cohort", "weeks", "disorder_group")] %>%
  unique() %>%
  group_by(weeks, disorder_group) %>% tally() %>%
  mutate(n = ifelse( n > obfuscation | isFALSE(obfuscation), n, 0.5), siteid = site)

patient_count_per_week_disorder_group %>%
  mutate(weeks = as.Date(weeks), 
         period = if_else(weeks < pandemic_start_date,
                          "Before the Start of the Pandemic",
                          "After the Start of the Pandemic"
         )
  ) %>% 
  ggplot(aes(x = as.Date(weeks), y= n, fill = period, group = disorder_group )) +
  geom_bar(stat = "identity")+
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  #geom_vline(xintercept = as.Date(late_pandemic_start_date),
  #           linetype = "dashed") +
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se= TRUE, aes(colour=period), colour= "black", size = 0.5) + 
  labs(x = "Week",
       y = "Patient Counts",
       title = "Weekly patient counts with mental health related ICD codes", 
       subtitle = "Grouped by mental health ICD subcategories") + 
  ylim(0, NA) + 
  facet_wrap(. ~ disorder_group, scales = "free_y") +
  theme(strip.text.x = element_text(size = 4),
        axis.text.x = element_text(size = 5, angle = 90), 
        axis.text.y = element_text(size = 6), 
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        title = element_text(size = 12),
        plot.subtitle = element_text(size = 11),
        legend.position = "none") +
  scale_fill_manual(values = cbPalette[-1])
ggsave("output/categorical_counts_inpatient_teens_all.png", width = 18, height = 9, units="cm")
```

>>>>

icdCategory <- read.delim("public-data/icd10Codes.txt", header = FALSE, colClasses = "character")
colnames(icdCategory) <- c("icd_code_letter", "icd_description_category")

all_codes <- obs_raw %>% 
  full_join( clinical_raw,  by = c("patient_num", "cohort", "days_since_admission")) %>%
  full_join( demo_raw, by = c("patient_num", "cohort")) %>% 
  mutate(icd_code_category = ifelse(concept_code %in% icdCodes$ICD10_Code, 
                               "psy", 
                               "others")) %>% 
  dplyr::filter( concept_type == "DIAG-ICD10", 
                 in_hospital == 1,
                 age < 18 & age > 10,
                 date >= history_start_date & date <= history_end_date) %>%
  dplyr::select( patient_num, cohort, sex, age, concept_type, concept_code, calendar_date,
                 in_hospital, n_hospitalisation, len_hospitalisation, date, weeks, month, year,
                 period, disorder_group, description, icd_code_category )

all_codes <- all_codes %>% 
  mutate( icd_code_letter = ifelse(substr(concept_code, 1, 1) %in% c("D", "H"), substr(concept_code, 1, 2), substr(concept_code, 1, 1) ))

all_codes <- left_join( all_codes, icdCategory) %>%
  mutate( icd_description_category = replace_na( ifelse( concept_code %in% icdCodes$ICD10_Code, "Psy", icd_description_category), "Others"))

@@

## Diagnoses count per week
Here we calculate the number of patients hospitalized per week (exactly, the number of patients being assigned a diagnosis on that week, hence usually corresponding to the end of hospitalization, ie not taking into account the length of the encounter). 
The barplot shows the weeks on the x-axis, and the patient counts on the y-axis. The horizontal lines is the (quadratic) temporal trend of these patient counts.

```{r}
patient_count_per_week_category <- all_codes[, c("patient_num", "cohort", "weeks", "icd_description_category")] %>%
  unique() %>%
  group_by(weeks, icd_description_category) %>%
  tally() %>%
  mutate(n = ifelse( n < obfuscation | isFALSE(obfuscation),
                     n,
                     0.5),
         siteid = site )

patient_count_per_week <- all_codes[, c("patient_num", "cohort", "weeks")] %>%
  unique() %>%
  group_by(weeks) %>%
  tally( name = "total_patients") %>%
  mutate(total_patients = ifelse( total_patients < obfuscation | isFALSE(obfuscation),
                     total_patients,
                     0.5),
         siteid = site )

patient_count_per_month_category <- all_codes[, c("patient_num", "cohort", "month", "icd_description_category")] %>%
  unique() %>%
  group_by(month, icd_description_category) %>%
  tally( name = "patients_per_category" ) %>%
  mutate(patients_per_category = ifelse( patients_per_category < obfuscation | isFALSE(obfuscation),
                     patients_per_category,
                     0.5),
         siteid = site )

patient_count_per_month <- all_codes[, c("patient_num", "cohort", "month")] %>%
  unique() %>%
  group_by(month) %>%
  tally( name = "total_patients") %>%
  mutate(total_patients = ifelse( total_patients < obfuscation | isFALSE(obfuscation),
                     total_patients,
                     0.5),
         siteid = site )

patient_count_monthly <- merge( patient_count_per_month_category, patient_count_per_month)
```

```{r}
#estimate the number of patients per week
patient_count_monthly %>%
  dplyr::filter(icd_description_category == "Psy") %>% 
  dplyr::mutate(month = as.Date(month)) %>% 
  dplyr::mutate(period = ifelse( month < pandemic_start_date,
                                  "before_pandemic",
                                 "during_pandemic"), 
                percentage = patients_per_category * 100 / total_patients) %>% 
  ggplot(aes(x = month, y = percentage, fill = period)) +
  #geom_line()+
  scale_fill_manual(values = cbPalette[c(3,4)]) +
  labs(y = "Percentage",
       x = "Calendar Months",
       title = "Percentage of patients with mental health related ICD codes (monthly)") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)

ggsave("output/overall_monthly_count_all.png", width = 18, height = 9, units="cm")
```

```{r}
patient_count_monthly %>%
  filter( icd_description_category %in% c("Psy")) %>%
  pivot_longer(names_to = "patients", cols=c(patients_per_category, total_patients)) %>%
  mutate( patients = ifelse( patients == "patients_per_category", "Psy", "Total")) %>%
  ggplot(aes(x = month, y = value, fill = patients, color = patients)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  scale_fill_manual(values = cbPalette) +
  scale_color_manual(values = cbPalette) +
  labs(y = "Counts",
       x = "Calendar Months",
       title = "Monthly patient counts with mental health related ICD codes") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)

ggsave("output/overall_weekly_count_all_comparison.png", width = 18, height = 9, units="cm")
```

```{r}
patient_count_monthly %>%
  filter( icd_description_category %in% c("Psy","Neoplams","Diseases of the respiratory system","Diseases of the digestive system")) %>%
  ggplot(aes(x = month, y = patients_per_category, fill = icd_description_category, color = icd_description_category)) +
  #geom_point(alpha=0.2) +
  geom_line() +
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  scale_fill_manual(values = cbPalette) +
  scale_color_manual(values = cbPalette) +
  labs(y = "Counts",
       x = "Calendar Months",
       title = "Months patient counts") +   
  theme(axis.text.x = element_text(angle=45, vjust = 1, hjust=1), legend.position = "bottom", legend.text = element_text(size=7), legend.title = element_text(size=7) ) +
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)

ggsave("output/overall_monthly_count_all_comparison_other_categories.png", width = 18, height = 9, units="cm")
```

@@

###percentage
patient_count_per_week_sex <- merge( patient_count_per_week_sex, patient_count_per_week )

patient_count_per_week_sex %>%
  mutate(percentage = n * 100 / total_patients ) %>%
  ggplot(aes(x = as.Date(weeks), y = percentage, fill = sex)) +
  geom_bar(stat="identity", position = "dodge")+
  scale_fill_manual(values = cbPalette) +
  labs(y = "Percentage",
       x = "Week",
       title = "Weekly patient percentage with mental health related ICD codes aggregated by sex",
       subtitle = "Per Sex") + 
  theme(legend.position = "bottom") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=sex), colour= "black", size = 0.5)
ggsave("output/overall_weekly_count_all.png", width = 18, height = 9, units="cm")

@@

patient_count_per_week_disorder_group <- input_data[, c("patient_num", "cohort", "weeks", "disorder_group")] %>%
  unique() %>%
  group_by(weeks, disorder_group) %>% tally( name = "psychiatric_patient_count") %>%
  mutate(psychiatric_patient_count = ifelse( psychiatric_patient_count > obfuscation | isFALSE(obfuscation), psychiatric_patient_count, 0.5), siteid = site)

patient_count_per_week_disorder_group <- merge( patient_count_per_week_disorder_group, patient_count_per_week)

patient_count_per_week_disorder_group %>%
  mutate(weeks = as.Date(weeks), 
         period = if_else(weeks < pandemic_start_date,
                          "Before the Start of the Pandemic",
                          "After the Start of the Pandemic"
         ), 
         percentage = psychiatric_patient_count * 100 / total_patients
  ) %>% 
  ggplot(aes(x = as.Date(weeks), y= percentage, fill = period, group = disorder_group )) +
  geom_bar(stat = "identity")+
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  #geom_vline(xintercept = as.Date(late_pandemic_start_date),
  #           linetype = "dashed") +
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se= TRUE, aes(colour=period), colour= "black", size = 0.5) + 
  labs(x = "Week",
       y = "Patient Percentage",
       title = "Weekly patient percentage with mental health related ICD codes", 
       subtitle = "Grouped by mental health ICD subcategories") + 
  ylim(0, NA) + 
  facet_wrap(. ~ disorder_group, scales = "free_y") +
  theme(strip.text.x = element_text(size = 4),
        axis.text.x = element_text(size = 5, angle = 90), 
        axis.text.y = element_text(size = 6), 
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        title = element_text(size = 12),
        plot.subtitle = element_text(size = 11),
        legend.position = "none") +
  scale_fill_manual(values = cbPalette[-1])
ggsave("output/categorical_counts_inpatient_teens_all.png", width = 18, height = 9, units="cm")
```
